\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2026

% The authors should use one of these tracks.
% Before accepting by the NeurIPS conference, select one of the options below.
% 0. "default" for submission
\usepackage{neurips_2026}
% the "default" option is equal to the "main" option, which is used for the Main Track with double-blind reviewing.
% 1. "main" option is used for the Main Track
%  \usepackage[main]{neurips_2026}
% 2. "position" option is used for the Position Paper Track
%  \usepackage[position]{neurips_2026}
% 3. "eandd" option is used for the Evaluations & Datasets Track
 % \usepackage[eandd]{neurips_2026}
 % if you need to opt-in for a single-blind submission in the E&D track:
 %\usepackage[eandd, nonanonymous]{neurips_2026}
% 4. "creativeai" option is used for the Creative AI Track
%  \usepackage[creativeai]{neurips_2026}
% 5. "sglblindworkshop" option is used for the Workshop with single-blind reviewing
 % \usepackage[sglblindworkshop]{neurips_2026}
% 6. "dblblindworkshop" option is used for the Workshop with double-blind reviewing
%  \usepackage[dblblindworkshop]{neurips_2026}

% After being accepted, the authors should add "final" behind the track to compile a camera-ready version.
% 1. Main Track
 % \usepackage[main, final]{neurips_2026}
% 2. Position Paper Track
%  \usepackage[position, final]{neurips_2026}
% 3. Evaluations & Datasets Track
 % \usepackage[eandd, final]{neurips_2026}
% 4. Creative AI Track
%  \usepackage[creativeai, final]{neurips_2026}
% 5. Workshop with single-blind reviewing
%  \usepackage[sglblindworkshop, final]{neurips_2026}
% 6. Workshop with double-blind reviewing
%  \usepackage[dblblindworkshop, final]{neurips_2026}
% Note. For the workshop paper template, both \title{} and \workshoptitle{} are required, with the former indicating the paper title shown in the title and the latter indicating the workshop title displayed in the footnote.
% For workshops (5., 6.), the authors should add the name of the workshop, "\workshoptitle" command is used to set the workshop title.
% \workshoptitle{WORKSHOP TITLE}

% Alternate titles:
% AD-MMTPP: Multimodal Temporal Point Processes for Alzheimer's Trajectory Prediction
% A Multimodal Temporal Point Process for Alzheimer's Disease and Related Dementias
% Language-Model Temporal Point Processes for Multimodal ADRD Prediction

% "preprint" option is used for arXiv or other preprint submissions
 % \usepackage[preprint]{neurips_2026}

% to avoid loading the natbib package, add option nonatbib:
%    \usepackage[nonatbib]{neurips_2026}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsmath}        % align, equation environments
\usepackage{amssymb}        % \mathbb, \mathcal
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{dsfont}         % \mathds{1} for indicator functions
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors
\usepackage{enumitem}       % compact lists

\newcommand{\soe}{\texttt{<SOE>}}
\newcommand{\eoe}{\texttt{<EOE>}}
\newcommand{\bos}{\texttt{<BOS>}}
\newcommand{\eos}{\texttt{<EOS>}}
\newcommand{\ttoken}[1]{\texttt{<#1>}}
\newcommand{\Hcal}{\mathcal{H}}

% Note. For the workshop paper template, both \title{} and \workshoptitle{} are required, with the former indicating the paper title shown in the title and the latter indicating the workshop title displayed in the footnote.
\title{Learning Generative Multimodal Trajectories of Alzheimer's Disease and Related Dementias}


% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.


\author{%
  % David S.~Hippocampus\thanks{Use footnote for providing further information
  %   about author (webpage, alternative address)---\emph{not} for acknowledging
  %   funding agencies.} \\
  % Department of Computer Science\\
  % Cranberry-Lemon University\\
  % Pittsburgh, PA 15213 \\
  % \texttt{hippo@cs.cranberry-lemon.edu} \\
  % examples of more authors
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}


\begin{document}


\maketitle


\begin{abstract}

\end{abstract}



\section{Introduction}
\label{sec:intro}

Alzheimer's disease and related dementias (ADRD) develop over many years, and the most informative measurements change with disease stage~\citep{self2023emerging}.
Molecular and fluid biomarkers tend to be most informative in early stages, while structural imaging, cognitive testing, and functional assessments become more informative later on.
ADRD also differs widely across patients, with subtypes that vary in clinical presentation, progression pattern, and rate of decline~\citep{young2018uncovering,fereshtehnejad2015new}.
Predictive models therefore have to handle both shifts over time in which modalities matter most and large variability between patients.

In ADRD, the prognostically informative signal is the joint evolution of cognitive scores, fluid and imaging biomarkers, and clinical events across visits, so a multimodal predictive model needs to represent the patient's trajectory rather than a static summary of it. Most current pipelines do the opposite. Cross-sectional benchmarks such as TADPOLE train a separate model per outcome from a fixed historical window, for example diagnosis at 24 months or ADAS-Cog at 36 months~\citep{marinescu2019tadpole}. Recent multimodal agentic pipelines fuse notes, imaging, and structured EHR through per-modality agents, but the EHR agent reduces each patient's visit history to a single feature vector, for example by max-pooling features over visits, before fitting the downstream classification or survival head~\citep{liu2026multidisciplinary}. The multimodal fusion is rich, but the temporal structure of the record is discarded before the label is predicted.
Two adjacent lines of work already know how to keep irregular time in the representation.
Marked temporal point processes (MTPPs), including neural and LLM-coupled variants, give density models for irregularly timed event streams~\citep{kong2025language,liu2025tpp,li2026longrange,berghaus2026incontext}.
Autoregressive transformers trained on tokenized electronic health records support arbitrary-horizon forecasting from a shared sequence backbone~\citep{renc2024zero,renc2025foundation}.
We combine them for the ADRD application and study whether a \emph{generative representation} of the multimodal trajectory, learned by next-token prediction on a language-model backbone, is useful on this problem.

We treat each patient history as a realization of a marked temporal point process and learn a multimodal representation of it by next-token prediction on a pretrained language-model backbone.
Each visit is serialized as a modality-tagged, delimited token sequence, and the backbone is adapted to it with parameter-efficient fine-tuning.
Next-token cross-entropy on this serialization corresponds to a discretized MTPP negative log-likelihood up to tokenizer resolution, so the backbone's hidden state becomes a learned multimodal event representation under a generative point-process objective.
The same representation supports diagnosis classification, biomarker regression, inter-visit time prediction, and multi-step trajectory generation through the same next-token decoding step, with no task-specific backbone and no collapse of the time axis.
Within this setup we compare three tokenization granularities, visit-level, modality-level, and atomic-event, to map the trade-off between context length and per-event supervision density, and we study a set of auxiliary heads (Hawkes intensity, Cox partial hazard, ordinal diagnosis) as different ways to shape the hidden state. The auxiliary heads supervise the representation and are not used for decoding.

\paragraph{Contributions.}
\begin{itemize}[leftmargin=*,itemsep=2pt,topsep=2pt]
    \item A generative representation-learning setup for multimodal ADRD trajectories in which a pretrained LLM backbone serves as a shared multimodal encoder under a point-process objective, with a formal link from token-level cross-entropy to the MTPP log-likelihood (§\ref{sec:problem}).
    \item Three tokenization variants, visit, modality, and atomic-event, that trade context length against per-event supervision density at an otherwise identical backbone, adaptation, and training objective (§\ref{sec:tokenization}).
    \item A suite of auxiliary heads (Hawkes NLL, Cox partial likelihood, ordinal regression) on the backbone's hidden states (§\ref{sec:regularization}) as different ways to shape the learned representation, evaluated against a pure next-token baseline.
    \item An empirical study on two ADRD cohorts, the TADPOLE subset of ADNI~\citep{adni,marinescu2019tadpole} and the tertiary-care EHR dataset EDUCATE, jointly ablating tokenization granularity and the auxiliary-head choice at fixed backbone, adaptation, and training recipe.
\end{itemize}

\section{Related Work}
\label{sec:related}

Five lines of prior work frame our approach: multimodal dementia prediction pipelines, neural temporal point processes coupled to language models, LLM fine-tuning on natural-language ``pseudo-notes'' rendered from structured EHR, non-language EHR tokenizers that train sequence models from scratch on custom event vocabularies, and EHR foundation-model encoders and generators pretrained directly on structured codes.
We treat them in turn and state where our variants sit relative to each.

\subsection{Multimodal Dementia Prediction}

Multimodal pipelines for ADRD tend to fuse across modalities while compressing the time axis of each patient's record. The TADPOLE challenge~\citep{marinescu2019tadpole} trains a separate model per outcome and fixed horizon from a hand-chosen historical window, and subsequent multimodal pipelines keep this style while adding modalities. Recent agentic pipelines~\citep{liu2026multidisciplinary} run per-modality agents over EHR, clinical notes, and imaging that each emit a prediction and supporting evidence, and aggregate them through a summary agent. The EHR agent there, however, collapses a patient's visit history to a single feature vector by max-pooling over visits before fitting a classification or survival head, so the multimodal fusion is rich but the temporal structure of the record is not propagated into the prediction. In this paper we study whether a generative representation-learning model that keeps the per-patient trajectory in the representation is useful for ADRD prediction. The backbone ingests the full token stream of the visit history under a point-process objective, and every prediction is read off the same representation by next-token decoding.

\subsection{Neural and LLM-coupled TPPs}

Marked temporal point processes have long been the standard tool for irregular event streams, with neural parameterizations of the conditional intensity developed over the last decade.
More recently, several works have tied TPPs to language-model backbones.
LanguageTPP~\citep{kong2025language} and TPP-LLM~\citep{liu2025tpp} cast next-event prediction as next-token prediction over a serialized event stream, using an LLM either to emit event distributions directly or to feed an intensity head on top of its hidden states.
MM-TPP~\citep{li2026longrange} extends this line to multi-modal marks, and OpenFIM~\citep{berghaus2026incontext} proposes a foundation intensity model trained on a broad mixture of heterogeneous event streams.
Our work sits in the LLM-coupled family but targets the ADRD application and is distinguished by (i) a controlled ablation over \emph{tokenization granularity} at fixed backbone, adaptation, and training objective, and (ii) treatment of auxiliary heads (intensity, hazard, ordinal) as representation-shaping objectives on the hidden state (§\ref{sec:regularization}) rather than primary outputs, so that the same model answers both discrete and time-valued tasks by next-token prediction.

\subsection{LLM Fine-Tuning on EHR Pseudo-Notes}

A parallel line of work renders tabular EHR records into natural-language-like sequences (``pseudo-notes'') and fine-tunes a general-purpose LLM on the resulting text.
MEME~\citep{lee2025clinical} is representative: patient events are templated into prose and the LLM is trained to predict downstream clinical outcomes.
Pseudo-note approaches inherit the linguistic priors of the backbone but pay a cost in numeric fidelity (scalar biomarkers become loosely formatted text) and in point-process structure (event boundaries and inter-event times are diluted into prose).
We keep events as explicitly delimited, modality-typed token strings with dedicated time tokens, which (a) preserves a clean event grid for the MTPP log-likelihood surrogate of Section~\ref{sec:problem} and (b) exposes per-event supervision to the loss.

\subsection{Non-Language EHR Tokenizers}

Independent of LLMs, several works tokenize clinical events into a purpose-built vocabulary and train a transformer from scratch on the resulting stream.
ETHOS~\citep{renc2024zero} encodes patient timelines into a discrete ``health language'' and trains a generative transformer for outcome prediction at arbitrary horizons. ARES~\citep{renc2025foundation} extends this generative-EHR line and is evaluated on similar prediction tasks.
These models share our autoregressive view of a patient record but forgo the linguistic priors of a pretrained LLM and re-learn tokenization, model, and task heads together.
Our variants retain the pretrained LLM via 4-bit quantization and LoRA adapters, so the only trainable structure specific to EHR is the vocabulary extension and the (optional) auxiliary heads.

\subsection{EHR Foundation-Model Encoders and Generators}

A parallel thread pretrains directly on structured EHR.
Masked- and next-code encoders such as BEHRT~\citep{li2020behrt} and Med-BERT~\citep{rasmy2021med} adapt BERT-style pretraining to ICD and diagnosis-code sequences, and CLMBR~\citep{steinberg2021language} casts representation learning on EHR as next-code language modeling over large structured corpora.
On the generative side, Foresight~\citep{kraljevic2024foresight} trains a GPT-style transformer on patient timelines for open-ended trajectory forecasting, conceptually close to ETHOS and ARES.
These four share our autoregressive or masked-LM view of a patient record but differ from our setup in three ways: they pretrain the tokenizer and the model together from scratch on structured codes, they do not expose a point-process log-likelihood surrogate, and they do not keep the linguistic priors of a general-purpose pretrained LLM.

\section{Method}
\label{sec:method}

We treat each patient's longitudinal record as a realization of a generative, marked temporal point process (MTPP) and learn it by autoregressive token prediction with a language-model backbone.
Section~\ref{sec:problem} develops the general formulation and prediction tasks.
Section~\ref{sec:tokenization} describes three concrete ways to implement this formulation as token sequences, and Sections~\ref{sec:model}--\ref{sec:training} specify the shared backbone and two-stage training recipe used for all three.
Section~\ref{sec:regularization} introduces the optional auxiliary heads (Hawkes, Cox, ordinal) attached to the backbone's representation.
Datasets, feature schema, and experimental protocol are described in Section~\ref{sec:experiments}.

\subsection{Problem Formulation}
\label{sec:problem}

Let $\mathcal{S} = \{(t_k, \mathbf{x}_k)\}_{k=1}^{K}$ denote a patient record with $K$ observations, where $t_k \in \mathbb{R}_{\geq 0}$ is the time since baseline (months) and $\mathbf{x}_k$ is a heterogeneous, possibly partially observed mark comprising diagnostic labels, cognitive scores, neuroimaging summaries, and structured codes.
Time-invariant covariates (sex, years of education, APOE-$\varepsilon$4 allele count) are collected in a context vector $\mathbf{c}$. Age is treated as a per-visit feature.
We write $\Hcal_k = \{(\mathbf{c}, (t_1, \mathbf{x}_1), \ldots, (t_k, \mathbf{x}_k))\}$ for the history up to and including the $k$-th observation and $\tau_k = t_k - t_{k-1}$ for the inter-observation interval.

A marked temporal point process is characterized by the conditional density of the next event given the history~\citep{daley2003introduction,shchur2021neural},
\begin{equation}
    p^*(t_{k+1}, \mathbf{x}_{k+1} \mid \Hcal_k)
    \;=\; \lambda^*(t_{k+1}, \mathbf{x}_{k+1}) \,
          \exp\!\Bigl(-\!\!\int_{t_k}^{t_{k+1}}\!\!\int \lambda^*(s, \mathbf{x})\,d\mathbf{x}\,ds\Bigr),
    \label{eq:mtpp}
\end{equation}
with conditional intensity $\lambda^*(\cdot \mid \Hcal_k)$.
Iterating the chain rule, the joint density of a full record factorizes into per-event conditionals,
\begin{equation}
    p^*\!\bigl(\{(\tau_k, \mathbf{x}_k)\}_{k=1}^{K} \,\big|\, \mathbf{c}\bigr)
    \;=\; \prod_{k=1}^{K} p^*(\tau_k, \mathbf{x}_k \mid \Hcal_{k-1}),
    \label{eq:mtpp_joint}
\end{equation}
so the MTPP negative log-likelihood of a record is the sum of the per-event log-densities of Eq.~\eqref{eq:mtpp}.

Rather than parameterize $\lambda^*$ directly, we serialize the record into a token sequence via a deterministic, delimited map
\[
    \phi \colon \bigl(\mathbf{c},\, \{(\tau_k, \mathbf{x}_k)\}_{k=1}^K\bigr)
    \;\longmapsto\;
    s_{1:L} \;=\; \bigl(s^{(c)},\, s^{(1)},\, \ldots,\, s^{(K)}\bigr),
\]
where $s^{(c)}$ encodes the static context and each event block $s^{(k)}$ encodes the inter-event time $\tau_k$ and mark $\mathbf{x}_k$ of visit $k$ (Section~\ref{sec:tokenization}).
An autoregressive language model factorizes $s_{1:L}$ by the chain rule, which we regroup by event blocks:
\begin{equation}
    p_\theta(s_{1:L})
    \;=\; \prod_{\ell=1}^{L} p_\theta(s_\ell \mid s_{<\ell})
    \;=\; p_\theta\!\bigl(s^{(c)}\bigr) \prod_{k=1}^{K} p_\theta\!\bigl(s^{(k)} \,\big|\, s^{(c)}, s^{(<k)}\bigr).
    \label{eq:ar}
\end{equation}
Because $\phi$ is injective on its image, each event-block factor $p_\theta\!\bigl(s^{(k)} \mid s^{(c)}, s^{(<k)}\bigr)$ is a parametric surrogate for the MTPP conditional $p^*(\tau_k, \mathbf{x}_k \mid \Hcal_{k-1})$, with continuous marks resolved to the tokenizer's numeric precision.
Token-level cross-entropy on $s_{1:L}$ therefore minimizes a discretized negative log-likelihood of Eq.~\eqref{eq:mtpp_joint}~\citep{liu2025tpp,kong2025language}.
This reformulation (i)~reuses a pretrained language-model backbone in place of a bespoke intensity network, (ii)~handles heterogeneous marks (categorical, ordinal, and continuous) under a single objective, and (iii)~unifies classification, regression, and forecasting as conditional next-token prediction.

\subsection{Three Granularity Variants}
\label{sec:tokenization}

All variants share a structural template for each patient:
\begin{equation}
    \bos \;\; \mathbf{c} \;\;
    \underbrace{\soe \; T(\tau_k) \; E(\mathbf{x}_k) \; \eoe}_{\text{event } k} \;\;
    \ldots \;\;
    \eos,
    \label{eq:template}
\end{equation}
where \soe/\eoe\ are abstract event delimiters, not learned vocabulary items: in implementation they are realized as newline boundaries plus the encoding-specific event prefix that opens each $E(\mathbf{x}_k)$ block (an event index $e_k\mid$ for (V), a modality tag $\langle g\rangle\mid$ for (M), a feature tag $\langle\text{feat}\rangle\mid$ for (A)).
The tokenizer already treats the newline as a separator, so no custom delimiter tokens are added to the vocabulary. The only added items are the modality/feature tags and the task prompts ($\ttoken{TASK\_*}$, $\ttoken{h}$, etc.).
The three variants differ in how the event payload $E(\mathbf{x}_k)$ is decomposed and, as a secondary consequence, in which continuous covariates can cleanly enter an event (the modality-level and atomic-event encodings restrict events to measurements with a natural modality/feature tag, so quantities such as age and raw intracranial volume are dropped in those two variants).
The three variants trade off context length against the density of supervision (fraction of tokens under a cross-entropy loss whose prefix varies across visits).
Let $F$ denote the number of observed feature dimensions at visit $k$ and $G$ the number of modality groups. For TADPOLE, $G$ covers imaging, cognitive, and diagnosis. For EDUCATE, $G$ covers diagnosis, lab, medication, and procedure.
Figure~\ref{fig:granularity} shows the same two-visit history under each variant.

\begin{figure}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{3pt}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{@{}p{0.28\linewidth} p{0.30\linewidth} p{0.38\linewidth}@{}}
\toprule
\textbf{(V) Visit-level} & \textbf{(M) Modality-level} & \textbf{(A) Atomic-event} \\
1 event / visit & $G$ events / visit & $F$ events / visit \\
\midrule
\ttfamily
\bos\ sex=F,educ=16,apoe4=0\newline
\soe\ \ttoken{t=0m}\newline
age=72\textbar vent=0.023\textbar hipp=0.004\textbar\newline
wb=0.78\textbar icv\_ml=1500\textbar\newline
dx=CN\ \eoe\newline
\soe\ \ttoken{t=12m}\newline
age=73\textbar vent=0.024\textbar hipp=0.004\textbar\newline
wb=0.77\textbar icv\_ml=1500\textbar\newline
mmse=27\textbar adas13=15\textbar dx=MCI\ \eoe\newline
\ttoken{TASK\_DX}\ \ttoken{h=12m}\newline
\rmfamily\textit{\scriptsize\,-- target --}\newline
\ttfamily AD
&
\ttfamily
\bos\ sex=F,educ=16,apoe4=0\newline
\soe\ \ttoken{t=0m}\ \ttoken{imaging}\newline
vent=0.023\textbar hipp=0.004\textbar wb=0.78\ \eoe\newline
\soe\ \ttoken{t=0m}\ \ttoken{diagnosis}\ CN\ \eoe\newline
\soe\ \ttoken{t=12m}\ \ttoken{imaging}\newline
vent=0.024\textbar hipp=0.004\textbar wb=0.77\ \eoe\newline
\soe\ \ttoken{t=12m}\ \ttoken{cognitive}\newline
mmse=27\textbar adas13=15\ \eoe\newline
\soe\ \ttoken{t=12m}\ \ttoken{diagnosis}\ MCI\ \eoe\newline
\ttoken{TASK\_DX}\ \ttoken{h=12m}\newline
\rmfamily\textit{\scriptsize\,-- target --}\newline
\ttfamily AD
&
\ttfamily
\bos\ sex=F,educ=16,apoe4=0\newline
\soe\ \ttoken{t=0m}\ \ttoken{vent}\ 0.023\ \eoe\newline
\soe\ \ttoken{t=0m}\ \ttoken{hipp}\ 0.004\ \eoe\newline
\soe\ \ttoken{t=0m}\ \ttoken{wb}\ 0.78\ \eoe\newline
\soe\ \ttoken{t=0m}\ \ttoken{dx}\ CN\ \eoe\newline
\soe\ \ttoken{t=12m}\ \ttoken{vent}\ 0.024\ \eoe\newline
\soe\ \ttoken{t=12m}\ \ttoken{hipp}\ 0.004\ \eoe\newline
\soe\ \ttoken{t=12m}\ \ttoken{wb}\ 0.77\ \eoe\newline
\soe\ \ttoken{t=12m}\ \ttoken{mmse}\ 27\ \eoe\newline
\soe\ \ttoken{t=12m}\ \ttoken{adas13}\ 15\ \eoe\newline
\soe\ \ttoken{t=12m}\ \ttoken{dx}\ MCI\ \eoe\newline
\ttoken{TASK\_DX}\ \ttoken{h=12m}\newline
\rmfamily\textit{\scriptsize\,-- target --}\newline
\ttfamily AD
\\
\midrule
\multicolumn{1}{@{}l}{\textit{shortest context}} &
\multicolumn{1}{l}{\textit{native block missingness}} &
\multicolumn{1}{l@{}}{\textit{densest supervision}} \\
\bottomrule
\end{tabular}
\caption{Two-visit patient history under the three granularities, with static context (sex, education, APOE-$\varepsilon$4) emitted once and MRI volumes reported as ratios to intracranial volume. Visit~1 at $t{=}0$ has no cognitive scores. Modality-level (M) drops the \ttoken{cognitive} sub-event, while (V) and (A) drop the missing features in place. The (V) encoding additionally emits current age and ICV (in mL) per visit. These are omitted by the modality-grouped (M) and atomic (A) encodings, which restrict events to measurements with a natural modality/feature tag. All three share the template of Eq.~\eqref{eq:template} and differ in how $E(\mathbf{x}_k)$ is decomposed and which fields it covers. The trailing task prompt \ttoken{TASK\_DX}\,\ttoken{h=12m} and the target response (here \texttt{AD}) are used only in Stage~2 supervised fine-tuning (\S\ref{sec:training}), where the cross-entropy loss is restricted to the response tokens. Stage~1 continued pretraining uses only the history portion, ending at the final \eoe\ of the last observed visit.}
\label{fig:granularity}
\end{figure}

\paragraph{(V) Visit-level encoding.}
One event per visit.
$E(\mathbf{x}_k)$ is a single row containing all observed feature--value pairs for that visit.
This is the most compact representation: a 10-visit TADPOLE patient uses roughly 10 rows, fits comfortably in a 4{,}096-token window, and preserves all same-visit correlations at the cost of coarse per-event supervision.

\paragraph{(M) Modality-level encoding.}
One event per modality group observed at the visit.
$E(\mathbf{x}_k)$ is split into $G$ sub-rows sharing a timestamp, each carrying only the features of its group (e.g., an imaging sub-event with MRI summaries, a cognitive sub-event with MMSE/CDRSB/ADAS).
Missing modalities are omitted rather than flagged with a missing-value token. A visit without a cognitive battery emits no cognitive sub-event.

\paragraph{(A) Atomic-event encoding.}
One event per individual measurement.
$E(\mathbf{x}_k)$ is decomposed into $F$ single-feature events of the form \ttoken{feat}$\;|\;t\;|\;v$.
This gives the finest per-token supervision (one loss boundary per measurement) and the cleanest treatment of feature-wise missingness, at the cost of $F{-}1$ extra timestamp repetitions per visit.

Numerical values are emitted as text at the backbone tokenizer's native resolution. Categorical labels (diagnosis, ICD/SNOMED codes) use a small set of added vocabulary items that we insert into the tokenizer and initialize from the mean of existing embeddings.
Task prompts are appended after the history as $\ttoken{TASK\_*} \, \ttoken{horizon}$ so a single model answers all five tasks and multiple horizons from a shared backbone.

\subsection{Model and Adaptation}
\label{sec:model}

We use a causal language model $f_\theta$ as the backbone.
Unless stated otherwise, $f_\theta$ is Qwen3.5-0.8B, loaded in 4-bit NormalFloat (NF4) quantization. Scale-up to Qwen3.5-2B is reserved for follow-up and reported only when stated.
For variants that introduce new tokens (category labels, modality tags, horizon bins), we extend the tokenizer and resize the embedding and output-projection matrices. New rows are initialized to the mean of the pretrained rows.

Trainable parameters are restricted to (i)~Low-Rank Adaptation \citep{hu2022lora} inserted in the attention and MLP projections ($\texttt{q,k,v,o\_proj}$, $\texttt{gate,up,down\_proj}$) as $W = W_0 + (\alpha/r)\, B A$ with $r=16$, $\alpha=64$, dropout $0.1$, and (ii)~the new token rows and the language-model head.
This reaches roughly $0.5\%$ of the backbone's parameters and fits on a single A100 80\,GB.

\subsection{Two-Stage Training}
\label{sec:training}

All three variants follow the same two-stage recipe, differing only in the token stream produced by their $E(\cdot)$.

\textbf{Stage~1 (continued pretraining).}
For each patient with $K$ observations we emit every prefix of length $2,\ldots,K$ as an independent sequence (sub-sequence augmentation), which expands $\sim$1{,}700 TADPOLE patients into $\sim$10{,}000 sequences.
The objective is next-token cross-entropy over the full sequence:
\begin{equation}
    \mathcal{L}_{\text{S1}}(\theta)
    \;=\; -\frac{1}{L} \sum_{\ell=1}^{L} \log p_\theta(s_\ell \mid s_{<\ell}).
    \label{eq:loss_s1}
\end{equation}

\textbf{Stage~2 (task supervised fine-tuning).}
For each valid split point we construct prompt--response pairs $(P, R)$ where $P$ is the tokenized history followed by $\ttoken{TASK\_*}$ and a horizon token, and $R$ is the target.
Examples across the five tasks are mixed.
The loss is computed only on the response tokens:
\begin{equation}
    \mathcal{L}_{\text{S2}}(\theta)
    \;=\; -\frac{1}{|R|} \sum_{\ell=1}^{|R|} \log p_\theta(r_\ell \mid P, r_{<\ell}).
    \label{eq:loss_s2}
\end{equation}
We study two regimes: \emph{next-visit} ($R$ targets only the immediate next observation) and \emph{horizon-conditioned} ($R$ targets any future observation whose offset matches the horizon token in $P$, so one model answers arbitrary horizons).
An optional augmentation regime applies modality, feature-value, and history dropout to the training sequences as a regularizer. Learning rate, batch size, hardware, augmentation rates, and per-stage epoch counts are deferred to Appendix~\ref{app:hparams}.

\subsection{Auxiliary-Head Objectives}
\label{sec:regularization}

The token-level objective of Eqs.~\eqref{eq:loss_s1}--\eqref{eq:loss_s2} treats inter-event time only through its surface form: the tokens of $T(\tau_k)$ appear in the loss exactly as any other symbol, and nothing forces the backbone's hidden states to encode time continuously or to respect the ordinal structure of diagnosis.
To probe which inductive biases help, we attach one \emph{optional} auxiliary head on the backbone representation during Stage~2 under the combined objective
\begin{equation}
    \mathcal{L} \;=\; \mathcal{L}_{\text{S2}} + \gamma\, \mathcal{L}_{\text{aux}},
    \label{eq:combined}
\end{equation}
with $\gamma$ linearly warmed from $0$ to $0.1$ over one epoch so the text objective is established before the head contributes gradients.
Heads are trained jointly with the text head but are not used at generation, and shape the hidden states only through back-propagation.

We consider three families. A \emph{Hawkes intensity head} fits a classical exponential-kernel Hawkes process under closed-form MTPP negative log-likelihood~\citep{hawkes1971spectra,ogata1981lewis,shchur2021neural,berghaus2026incontext}. A \emph{Cox proportional-hazards head} is trained under the Cox partial log-likelihood~\citep{cox1972regression} with Breslow ties~\citep{breslow1972contribution}, in two sub-variants: accumulated risk sets across mini-batches~\citep{katzman2018deepsurv,kvamme2019time}, and an attention-pooled history~\citep{lin2017structured}. An \emph{ordinal head} exploits the CN\,$<$\,MCI\,$<$\,AD order via cumulative binary cross-entropy~\citep{frank2001simple,niu2016ordinal}, carries no timing information, and tests whether an order-aware classification signal alone suffices to shape the hidden states.
Ablations compare each tokenization variant across these heads and a pure-text baseline ($\gamma=0$) at otherwise matched configurations. Full formulations are deferred to Appendix~\ref{app:aux_heads}.

\section{Experiments}
\label{sec:experiments}

\subsection{Experimental Setup}
\label{sec:setup}

Both datasets follow the same abstract schema $(t_k, \mathbf{x}_k)$ with $\mathbf{x}_k$ structured as modality-typed sub-vectors, so the three model variants of Section~\ref{sec:tokenization} apply uniformly. Dataset-specific feature lists are deferred to Appendix~\ref{app:datasets}.
Prediction tasks are defined per dataset, as the two sources differ in the targets available.

\paragraph{TADPOLE (ADNI).}
The TADPOLE challenge subset of ADNI \citep{adni,marinescu2019tadpole} provides $\sim$1{,}700 patients in a training split (D1) and a held-out rollover split (D2, later ADNI follow-up visits acquired after the challenge cut-off). Each patient carries static demographic covariates and a per-visit record of cognitive scores, MRI-derived volumes, and a CN/MCI/AD diagnosis label. The full schema, preprocessing recipe, and per-variant field lists are in Appendix~\ref{app:datasets}.
Given a history $\Hcal_k$ and horizon $h>0$, we define five prediction tasks on TADPOLE: diagnosis at $h$ (3-way CN/MCI/AD), ADAS-Cog at $h$, ventricle-to-ICV ratio at $h$, inter-visit interval $\tau_{k+1}$, and multi-step trajectory generation by ancestral sampling (drawing one event at a time from the model's conditional distribution and feeding each draw back as context for the next).
Experiments in this paper report diagnosis. The remaining four are reserved for future work.
We also consider a \emph{DX-hidden} ablation in which the diagnosis field is dropped from each visit event at training and inference, to measure how much of the TADPOLE signal passes through the cognitive and MRI fields alone.

\paragraph{Evaluation protocol.}
TADPOLE diagnosis is evaluated with 5-fold stratified group $k$-fold cross-validation on D1 (groups = patient IDs, seed~42), so that no patient appears in both train and validation.
Each fold runs the full two-stage recipe (§\ref{sec:training}) and scores the held-out patients by balanced class accuracy (BCA) over the three classes \{CN, MCI, AD\}. BCA is used as the primary metric because ADNI is strongly class-imbalanced and random-chance accuracy would otherwise dominate a plain accuracy score.
The D2 rollover split is held out as a secondary, temporally shifted test set.
The Stage-2 sweep covers 192 fine-tuned models per fold across tokenization, prompt style, DX visibility, auxiliary-head choice, augmentation, and target regime. The three-stage pipeline (pretrain, fine-tune, evaluate), the list of 12 modes, and the decomposition of the 48 mode-loss pairs are described in Appendix~\ref{app:sweep}.

\paragraph{EDUCATE.}
EDUCATE is a tertiary-care electronic health record dataset covering dementia and related disorders, with laboratory, medication, diagnosis, and procedure events carrying absolute timestamps and mapped to the OMOP Common Data Model (a standardized health-record vocabulary). The ontology crosswalk and full event schema are in Appendix~\ref{app:datasets}.
Four prediction tasks are defined on EDUCATE: binary healthy/non-healthy classification on the full history, five-way class (AD, MCI, frontotemporal dementia (FTD), dementia with Lewy bodies (DLB), other) restricted to non-healthy patients, inter-visit interval $\tau_{k+1}$, and multi-step trajectory generation.
Because diagnosis codes directly encode the class labels, the two classification tasks are additionally evaluated with a \emph{leak-safe} preprocess that removes all dementia/ADRD diagnostic codes and text mentions at load time (specific code prefixes in Appendix~\ref{app:datasets}). Gaps between the two preprocess variants quantify code-level leakage.

\paragraph{Decoding for diagnosis.}
For the diagnosis task reported here, we decode greedily at temperature $T_{\text{gen}}=0.05$ with constrained generation over $\{\text{CN},\text{MCI},\text{AD}\}$, conditioned on the tokenized history followed by the task prompt $\ttoken{TASK\_DX}\,\ttoken{h}$.
The same prompt template is reused across horizons so that a single trained model handles all horizon queries.

\subsection{Results}
\label{sec:results}

We study how the tokenization granularity of §\ref{sec:tokenization} and the auxiliary-head choices of §\ref{sec:regularization} affect TADPOLE diagnosis prediction.
All three encodings (V, M, A) share the same backbone, adapter, and two-stage training recipe, so any gap in BCA traces to how the event payload is encoded.
We then run each encoding under each of the auxiliary heads (pure text, Hawkes NLL, Cox partial likelihood with accumulated-risk-set and attention-pooled sub-variants, and ordinal cumulative BCE), at otherwise matched settings, to test which inductive bias on the hidden states transfers to next-token diagnosis accuracy.

\paragraph{Headline results on ADNI TADPOLE.}
[Placeholder: best configuration across tokenization variants and auxiliary heads on 3-way CN/MCI/AD diagnosis, reported as mean $\pm$ std BCA over 5 folds of D1 and on the D2 rollover split.]

\paragraph{Headline results on EDUCATE.}
[Placeholder: best configuration across tokenization variants and auxiliary heads on binary healthy/non-healthy classification and on 5-way AD/MCI/FTD/DLB/other classification, under the leak-safe preprocess.]

\paragraph{Tokenization granularity.}
[Placeholder: BCA for encodings V, M, and A at matched backbone, adapter, and training recipe, averaged over folds.]

\paragraph{Auxiliary-head ablation.}
[Placeholder: each encoding under pure text ($\gamma=0$), Hawkes NLL, Cox partial likelihood (plain / accumulated / attention-pooled), and ordinal cumulative BCE, all at $\gamma=0.1$, under the same splits and horizons.]

\paragraph{DX-hidden and augmentation ablations.}
[Placeholder: BCA for the best encoding under the DX-hidden preprocess and under the \emph{moderate} augmentation setting (Section~\ref{sec:training}), reported alongside the default DX-visible, no-augmentation configuration.]

\section{Conclusion}




\section*{References}
\bibliographystyle{plainnat}
\bibliography{references}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix

\section{Appendix}
\label{app:appendix}

\subsection{Auxiliary Heads}
\label{app:aux_heads}

This appendix gives the formal definition of each auxiliary head of §\ref{sec:regularization}, matched to the \texttt{brainx} implementation.
Notation: $h_k \in \mathbb{R}^d$ is the backbone's last hidden state at event position $k$. $t_k$ and $m_k$ are the event time and mark. $d_k \in \{0,1,2\}$ encodes diagnosis (CN, MCI, AD). $\sigma(\cdot)$ is the logistic function. $\mathcal{B}$ is a mini-batch of valid event positions.
The pure-text baseline sets $\mathcal{L}_{\text{aux}} = 0$ in Eq.~\eqref{eq:combined}, so Stage~2 optimizes only the response-token cross-entropy $\mathcal{L}_{\text{S2}}$ of Eq.~\eqref{eq:loss_s2}. The subsections below cover the five heads that specify a non-trivial $\mathcal{L}_{\text{aux}}$.

\subsubsection{Hawkes NLL (closed form)}
Small per-mark projections produce $(\mu_m^{(k)}, \alpha_m^{(k)}, \beta_m^{(k)}) = \mathrm{softplus}\bigl(W_{\{\mu,\alpha,\beta\},m}\, h_k\bigr)$ at every event position $k$, parameterizing a piecewise Hawkes intensity in the classical exponential-kernel form~\citep{hawkes1971spectra,shchur2021neural},
\begin{equation}
    \lambda_m(t \mid \Hcal_k)
    \;=\; \mu_m + \!\!\sum_{j:\, t_j < t,\; m_j = m}\!\! (\alpha_m - \mu_m)\, e^{-\beta_m (t - t_j)}.
    \label{eq:hawkes}
\end{equation}
The MTPP negative log-likelihood
\begin{equation}
    \mathcal{L}_{\text{NLL}}
    \;=\; \int_0^{T} \sum_m \lambda_m(s)\, ds \;-\; \sum_{k=1}^{K} \log \lambda_{m_k}(t_k)
    \label{eq:nll}
\end{equation}
is computed in closed form: the compensator admits
\begin{equation}
    \int_0^T \lambda_m(s)\, ds
    \;=\; \mu_m^{(K)}\, T
    \;+\; \!\!\sum_{j:\, m_j = m}\!\!
          \frac{\alpha_m^{(j)} - \mu_m^{(j)}}{\beta_m^{(j)}}
          \Bigl(1 - e^{-\beta_m^{(j)}(T - t_j)}\Bigr),
    \label{eq:hawkes_compensator}
\end{equation}
where $\mu_m^{(K)}$ is the baseline read off the last event~\citep{ogata1981lewis}.
The implementation follows the \texttt{PiecewiseHawkesIntensity} interface of OpenFIM~\citep{berghaus2026incontext}.

\subsubsection{Cox partial log-likelihood}
A scalar log-hazard $f(h_k) = w^\top h_k + b$ is read off the last hidden state at each event position, with $w \in \mathbb{R}^d$ and $b \in \mathbb{R}$.
Events marking an observed class transition are labeled $\delta_k = 1$, otherwise $\delta_k = 0$.
With times sorted descending and within-batch ties resolved by the Breslow approximation~\citep{breslow1972contribution,cox1972regression},
\begin{equation}
    \mathcal{L}_{\text{Cox}}
    \;=\; -\frac{1}{N_\delta}\!\! \sum_{k:\, \delta_k = 1}\!\!
    \Bigl[\, f(h_k)
    \;-\; \log \!\!\!\sum_{j \in \mathcal{R}(t_k)}\!\!\! \exp\!\bigl(f(h_j)\bigr) \,\Bigr],
    \qquad N_\delta = |\{k: \delta_k = 1\}|,
    \label{eq:cox_app}
\end{equation}
where the risk set $\mathcal{R}(t_k) = \{j : t_j \geq t_k\}$ is evaluated via a descending-time cumulative log-sum-exp.

\subsubsection{Cox with accumulated risk set}
At small mini-batch sizes the ranking signal in Eq.~\eqref{eq:cox_app} weakens.
Following~\citep{katzman2018deepsurv,kvamme2019time}, we buffer $\{f(h_k), t_k, \delta_k\}$ across $N_{\text{acc}} = 4$ consecutive mini-batches and compute Eq.~\eqref{eq:cox_app} on the combined buffer, which gives an effective risk set of $|\mathcal{B}| \cdot N_{\text{acc}}$.
Gradients flow only through the most recent mini-batch. Earlier entries are detached and provide ranking context.
The Cox loss is computed and back-propagated only at every $N_{\text{acc}}$-th step, when the buffer fills. On the intervening $N_{\text{acc}}{-}1$ steps only the text loss is applied.

\subsubsection{Cox with attention-pooled hidden state}
A learned query $q \in \mathbb{R}^d$ attends over the full history $\{h_1, \ldots, h_K\}$ via scaled dot-product attention,
\begin{equation}
    \alpha_k
    \;=\; \frac{\exp\bigl(\langle q, h_k \rangle / \sqrt{d}\bigr)}
               {\sum_j \exp\bigl(\langle q, h_j \rangle / \sqrt{d}\bigr)},
    \qquad
    \bar h
    \;=\; \sum_k \alpha_k\, h_k,
    \label{eq:attn_pool}
\end{equation}
after which the log-hazard is $f(\bar h) = w^\top \bar h + b$, following self-attentive pooling~\citep{lin2017structured}.
Eq.~\eqref{eq:cox_app} is then applied on the attention-pooled log-hazards.

\subsubsection{Ordinal cumulative BCE}
A two-logit head $g: h_k \mapsto (z_{k,1}, z_{k,2}) \in \mathbb{R}^2$ parameterizes cumulative probabilities
$P(d_k \geq 1 \mid h_k) = \sigma(z_{k,1})$ and $P(d_k \geq 2 \mid h_k) = \sigma(z_{k,2})$
under the cumulative-link ordinal formulation~\citep{frank2001simple,niu2016ordinal}:
\begin{equation}
    \mathcal{L}_{\text{ord}}
    \;=\; \frac{1}{2|\mathcal{B}|} \sum_{k \in \mathcal{B}}\!
    \Bigl[\,
      \mathrm{BCE}\!\bigl(z_{k,1},\, \mathds{1}[d_k \geq 1]\bigr)
    + \mathrm{BCE}\!\bigl(z_{k,2},\, \mathds{1}[d_k \geq 2]\bigr)
    \,\Bigr].
    \label{eq:ord}
\end{equation}
Posterior class probabilities used at decoding time are recovered as $P(\text{CN}) = 1 - \sigma(z_1)$, $P(\text{MCI}) = \sigma(z_1) - \sigma(z_2)$, $P(\text{AD}) = \sigma(z_2)$.

\subsection{Datasets and Preprocessing}
\label{app:datasets}

\subsubsection{TADPOLE (ADNI)}
Three static covariates $\mathbf{c}$ (sex, years of education, APOE-$\varepsilon$4 allele count) are emitted once per patient. Nine observation fields are available per visit: current age, diagnosis $d \in \{\text{CN}, \text{MCI}, \text{AD}\}$, three cognitive scores (MMSE, CDRSB, ADAS-13), three MRI volumes normalized as ratios to intracranial volume (ventricles, hippocampus, whole-brain), and ICV itself in milliliters. Time since baseline $t$ (months) and inter-visit interval $\Delta t$ are attached as tokens at each event position.
The (V) encoding emits all nine fields per visit. The (M) and (A) encodings drop age and ICV-in-mL (neither has a clean modality/feature tag) and emit the remaining seven grouped by modality or atomized by feature.
Preprocessing: raw MRI volumes are read from the ADNI FreeSurfer release, the three normalized volumes are computed as ratios to ICV, ICV is converted from mm$^3$ to mL, ADNI transition labels (e.g., \texttt{NL to MCI}, \texttt{MCI to Dementia}) are collapsed to the 3-way CN/MCI/AD taxonomy, and patients with fewer than two visits are dropped.

\subsubsection{EDUCATE}
Each patient carries an irregular stream of diagnosis, laboratory, medication, and procedure events with absolute timestamps. Source codes are mapped to the OMOP Common Data Model through a project-specific crosswalk before tokenization.
The leak-safe preprocess removes all dementia/ADRD ICD-10 prefixes (G30, G31, F00--F03, $\ldots$) and SNOMED/free-text mentions at load time, so no dementia code reaches the tokenizer.

\subsection{Sweep design and pipeline}
\label{app:sweep}

Each fold runs a three-stage pipeline. Stage~1 continues pretraining under plain next-token cross-entropy with no auxiliary head, producing one LM checkpoint per (mode, augmentation) pair. Stage~2 attaches the optional auxiliary head on top of this Stage-1 checkpoint and fine-tunes under the Stage-2 objective of Eq.~\eqref{eq:loss_s2}. Stage~3 evaluates each Stage-2 model on the held-out fold and on the D2 rollover split. Because auxiliary heads attach only in Stage~2, one Stage-1 checkpoint is shared across all compatible losses at the same mode and augmentation, so Stage-1 cost grows with the mode axis but not with the loss axis.

\subsubsection{Modes}

The Stage-1 grid covers 12 modes. Each mode fixes a tokenization granularity (V, M, or A, as defined in §\ref{sec:tokenization}), a prompt style, and a diagnosis-visibility setting:
\begin{itemize}[leftmargin=*,itemsep=2pt,topsep=2pt]
    \item \emph{Base} modes (DX visible, ChatML-style system prompt, loss over the full input sequence): V-base, M-base, A-base.
    \item \emph{Strict} modes (DX visible, structured prompt, loss restricted to response tokens per Eq.~\eqref{eq:loss_s2}): V-strict, M-strict, A-strict.
    \item \emph{No-DX} modes (DX dropped from history, base prompt): V-nodx, M-nodx, A-nodx.
    \item \emph{FIM} modes (strict prompt with summary temporal features prepended, paired only with the Hawkes intensity head): V-fim, M-fim, A-fim.
\end{itemize}
Running the 12 modes under 2 augmentation settings (none, moderate) gives 24 Stage-1 checkpoints per fold.

\subsubsection{Mode-loss pairs}

The 48 Stage-2 mode-loss pairs decompose as:
\begin{itemize}[leftmargin=*,itemsep=2pt,topsep=2pt]
    \item Base and strict modes (6) $\times$ 5 auxiliary-head settings (pure text, Cox, accumulated-risk Cox, attention-pooled Cox, ordinal) = 30.
    \item No-DX modes (3) $\times$ 5 auxiliary-head settings = 15.
    \item FIM modes (3) $\times$ 1 auxiliary-head setting (Hawkes) = 3.
\end{itemize}
Crossing the 48 pairs with 2 augmentation settings and 2 target regimes (next-visit, horizon-conditioned) gives 192 Stage-2 models per fold. Stage-3 evaluates each of these on the held-out fold and on the D2 rollover split.

\subsection{Training Hyperparameters}
\label{app:hparams}

Both stages use learning rate $10^{-4}$ and per-device batch size $4$ on a single A100 80\,GB GPU. Stage~1 runs for 7 epochs without augmentation and 9 epochs with augmentation. Stage~2 runs for 4 epochs without augmentation and 5 epochs with augmentation. When augmentation is enabled, it applies 10\% modality dropout, 2\% feature-value noise, and 15\% history dropout to the training sequences. For the auxiliary-head combined objective of Eq.~\eqref{eq:combined}, $\gamma$ is linearly warmed from $0$ to $0.1$ over one epoch.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\input{checklist.tex}


\end{document}